Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
paper: null made macro, since had null and NULL

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we instantiate \textit{Impute} with the CE-CART\cite{burgette2010multiple} algorithm, but we highlight that
our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no null values for
attributes in $C$.

@@

\subsection{Imputation placement}
\label{sec:placement}
Imputation operators must be placed so that no relational operator receives a tuple containing null in an attribute that the operator examines, regardless of the state of the data in the base tables.

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null. We compute a dirty set for each table using the table histograms, which track the number of null values in each column. The dirty set for a query plan can be computed recursively as follows:

@@

Note that \textsc{Dirty} over-approximates the set of attributes that contain null. For example, a filter might remove all tuples which contain null, but the dirty set would be unchanged. We choose to over-approximate to ensure that all null values are explicitly imputed or dropped.

@@

\srm{I find this table hard to follow -- I think it'd be good to explain one or two of the rows in detail, and to explain what the set L contains.}
Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). Imputing null fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

>>>>

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we instantiate \textit{Impute} with the CE-CART\cite{burgette2010multiple} algorithm, but we highlight that
our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

\subsection{Imputation placement}
\label{sec:placement}
Imputation operators must be placed so that no relational operator receives a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a \nullv{} value, there are cases where an imputation operator is required. To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain \nullv{}. We compute a dirty set for each table using the table histograms, which track the number of \nullv{} values in each column. The dirty set for a query plan can be computed recursively as follows:

@@

Note that \textsc{Dirty} over-approximates the set of attributes that contain \nullv{}. For example, a filter might remove all tuples which contain \nullv{}, but the dirty set would be unchanged. We choose to over-approximate to ensure that all \nullv{} values are explicitly imputed or dropped.

@@

\srm{I find this table hard to follow -- I think it'd be good to explain one or two of the rows in detail, and to explain what the set L contains.}
Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). Imputing \nullv{} fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

### 2017-03-01 paper/experiments.tex
paper: null made macro, since had null and NULL

The original tables have a large number of attributes, in some cases providing more granular tests results or alternative metrics.
We focused on a subset of the attributes for each table to simplify the presentation and exploration of queries.
\Cref{table:nhanes-description} shows the attributes selected, along with the percentage of null values for each attribute.
For readability, we have replaced the NHANES variable names with self-explanatory attribute names.

@@

We use a version of the data that has been pre-processed, but where missing values remain.
For example, 46.6\% of \textit{commutetime} responses are missing. However, it is worth
noting that some of the missing values are also expected, given the way the data has been
de-normalized. For example, \textit{bootcamploanyesno}, a binary attribute encoding whether
a respondent had a loan for a bootcamp, is expected to be null for participants who did not
attend a bootcamp.

>>>>

The original tables have a large number of attributes, in some cases providing more granular tests results or alternative metrics.
We focused on a subset of the attributes for each table to simplify the presentation and exploration of queries.
\Cref{table:nhanes-description} shows the attributes selected, along with the percentage of \nullv{} values for each attribute.
For readability, we have replaced the NHANES variable names with self-explanatory attribute names.

@@

We use a version of the data that has been pre-processed, but where missing values remain.
For example, 46.6\% of \textit{commutetime} responses are missing. However, it is worth
noting that some of the missing values are also expected, given the way the data has been
de-normalized. For example, \textit{bootcamploanyesno}, a binary attribute encoding whether
a respondent had a loan for a bootcamp, is expected to be \nullv{} for participants who did not
attend a bootcamp.

### 2017-03-01 paper/introduction.tex
paper: null made macro, since had null and NULL

Many databases have large numbers of missing or NULL values;  these can arise for a variety of reasons, including missing source data (e.g., unknown facts),
missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where NULL values are substituted for
bad values~\cite{kim2003}.  Such NULL values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
before processing is often desirable.

% Solution
One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we present a new system, called \ProjectName{}, which is the designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database.

@@

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain null values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
\item \textbf{Proof of optimality:}
 We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

Many databases have large numbers of missing or \nullv{} values;  these can arise for a variety of reasons, including missing source data (e.g., unknown facts),
missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where \nullv{} values are substituted for
bad values~\cite{kim2003}.  Such \nullv{} values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
before processing is often desirable.

% Solution
One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we present a new system, called \ProjectName{}, which is designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database.

@@

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

### 2017-03-01 paper/relatedwork.tex
paper: remove formal proof
paper: null made macro, since had null and NULL

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of nulls. As early as 1973,~\cite{codd1973understanding}
provides a treatment of the semantics of null. Multiple
papers have described various (at times conflicting) treatments
of nulls~\cite{grant1977null}. \ProjectName's main design invariant - no relational operator
sees missing values for attributes it must operate on and users should never see
missing data - eliminates
the need to handle null value semantics, while guaranteeing soundness (modulo
imputation strategy) of the query evaluation.

@@

Designers of data stream processing systems frequently confront missing values
and consider them carefully in query processing. Often, if sensor error is the
cause of missing values, values can be imputed with high confidence. In
\cite{fernandez2009inter}, feedback punctuation is used to dynamically
reconfigure the query plan for state-dependent
optimizations as data continue to arrive. One of the applications of this framework is to avoid
expensive imputations as real-time conditions change. 
\ProjectName{} focuses on handling
missing values, rather than providing a general framework for planning feedback. Furthermore,
it provides the ability
to trade imputation quality for runtime performance. We prove our
algorithm and cost model produces an optimal set of plans to choose from,
given the constraints of the search space.

>>>>

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of \nullv{}. As early as 1973,~\cite{codd1973understanding}
provides a treatment of the semantics of \nullv{}. Multiple
papers have described various (at times conflicting) treatments
of nulls~\cite{grant1977null}. \ProjectName's main design invariant - no relational operator
sees missing values for attributes it must operate on and users should never see
missing data - eliminates
the need to handle \nullv{} value semantics, while guaranteeing soundness (modulo
imputation strategy) of the query evaluation.

@@

Designers of data stream processing systems frequently confront missing values
and consider them carefully in query processing. Often, if sensor error is the
cause of missing values, values can be imputed with high confidence. In
\cite{fernandez2009inter}, feedback punctuation is used to dynamically
reconfigure the query plan for state-dependent
optimizations as data continue to arrive. One of the applications of this framework is to avoid
expensive imputations as real-time conditions change. 
\ProjectName{} focuses on handling
missing values, rather than providing a general framework for planning feedback. Furthermore,
it provides the ability
to trade imputation quality for runtime performance. 
%We prove our
%algorithm and cost model produces an optimal set of plans to choose from,
%given the constraints of the search space.
